---
title: "Core Project"
author: "Serge"
date: "7/23/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# JHU Reproducible Research, Assignment 1

Load Packages for use in this exercise
```{r}
library('lubridate')
library('dplyr')
```


Unzip and read in activity file
```{r}

data <- read.table(unzip('activity.zip'), header = TRUE, sep = ',')

```

Convert date column from factor to date
```{r}
data$date <- ymd(as.character(data$date))
```


#Exploratory DAta Analysis
```{r}
#How many days does this cover?  288 readings per day, 61 days, starting October 1, 2012, and ending November 30, 2012.
(data$date)

```

Instead of yielding an actual number of steps for each time segment, most entries are NA.  Specifically, out 17568 rows of data total, 13,318 are NA entries.

```{r}
nrow(data) - nrow(data %>% filter(steps != is.na(steps)))
```


#What are the mean total number of steps taken per day?

```{r}
totals <- data %>%
  group_by(date) %>%
  summarize(total = sum(steps))

# mean number of steps taken (ignoring days that include NA values)
round(mean(totals$total, na.rm = TRUE), 0)
```

```{r}
hist(data$date)
```

#Mean Daily Activity Pattern, with NA's removed from calculation of mean.

```{r}
timeseries <- data %>%
  group_by(interval) %>%
  summarize(mean = mean(steps, na.rm = TRUE))

with(timeseries, plot(interval, mean, type = 'l'))
```

